---
title: "Seminar 2 — Confidence Intervals & Effect Sizes"
subtitle: "PSY40760 Advanced Data Analysis and Statistics in R"
date: 2026-09-30
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: false
---

# Worksheet 2

## Welcome

This is the worksheet for our second session. It lists the reading and the to-do items for this week, then guides you through the practical.

## Session at a glance

|  |  |
|----|----|
| **Workshop Date** | Wednesday 30 September 2026 |
| **Due Date** | Sunday 11 October 2026, 23:59 |
| **Time** | 14:00 – 16:50 (170 min) |
| **Room** | L249-SUTH |
| **Staff** | Dr Ondrej Zika, Cameron McCabe |
| **Theory** | Confidence intervals & effect sizes |
| **Practical** | Paths, packages, the t-distribution; importing and wrangling data; CIs and effect sizes in R |

## Before you arrive

### 1. Set up the course folder

All practicals of the module live in **one course folder**, with the same structure on every computer. For the first two seminars it looks like this:

```         
psy40760-practicals-2026/          <- the course folder (the "root")
├── .root-practicals               <- empty marker file: marks the root (section 1.3)
├── .Rprofile                      <- renv set-up files (section 1.2)
├── renv.lock
├── renv/
├── seminar1-intro-to-stats-and-R/
│   ├── seminar-01-worksheet.qmd
│   └── Cheatsheets/
└── seminar-02-confidence-intervals-and-effect-sizes/
    ├── seminar-02-worksheet.qmd       <- this worksheet
    └── seminar-02-worksheet_files/    <- this week's data
        ├── weight_trial.csv
        ├── weight_followup_wide.csv
        └── simulate_datasets.R
```

The renv files and `.root-practicals` sit at the top of the course folder; each seminar has its own folder below it, holding the worksheet, its `_files` folder (data) and any extra materials. Keep the folder and file names exactly as shown: the code in the worksheets relies on them.

**Option A (easiest): clone the course repository.** Once, in a terminal:

```         
git clone git@github.com:ozika/psy40760-practicals-2026.git
```

Before each practical, run `git pull` inside the folder to get the new week's files. Then, in RStudio, point R at the course folder (*Session → Set Working Directory → Choose Directory…*) and run, in the console:

### Quick step-by-step guide

```{r}

# Cheatsheet
# tell R where your project lives
setwd("[put your path here]")

# install renv (you might alrady have it)
install.packages("renv")

#load renve
library(renv)

# install here
renv::install("here")

# load here
library(here)

# tell here where your home dir is 
here::i_am(".root-practicals") 

# check that this is correct
here::here(".root-practicals")

# load the environment
renv::load(here(".root-practicals"))

# install additional packages
renv::install("dplyr")

---
  
# Common issues 
#  Error: '\U' used without hex digits in character string (<input>:1:11)


```

```{r}
#| eval: false
renv::load()      # switch this R session to the course's package library
renv::restore()   # install this week's packages (section 1.2)
```

**Option B: set it up by hand.**

1.  Create a folder called `psy40760-practicals-2026` wherever you keep your work.
2.  In RStudio, point R at that folder (*Session → Set Working Directory → Choose Directory…*) and create the marker file (step 2 below).
3.  Download each seminar folder and place it inside the course folder exactly as in the tree above.
4.  Set up renv once and install this week's packages (step 4 below). Here the folder is your own project, so you install the packages yourself.

```{r}
#| eval: false
# Step 2: an empty marker file; it is hidden because the name starts with a dot
file.create(".root-practicals")

# Step 4: set up renv and install this week's packages
renv::init()                               # creates .Rprofile, renv.lock and renv/
renv::install(c("here", "dplyr", "tidyr"))
renv::snapshot()                           # records the versions in renv.lock
```

Either way, you do not need to create an RStudio Project: renv takes care of the packages, and the `.root-practicals` file takes care of finding the data.

### 2. DataCamp

This week's two DataCamp courses are listed under Part 2. Use the same UCD account as last week.

### 3. Reading

|  |  |
|----|----|
| **Theory** | *Statistical Thinking for the 21st Century* (Poldrack) — [Ch. 10, Quantifying effects and designing studies](https://statsthinking21.github.io/statsthinking21-core-site/ci-effect-size-power.html) |
|  | *Modern Statistics with R* — [Ch. 3, The cornerstones of statistics](https://www.modernstatisticswithr.com/basicstatistics.html) (especially §3.5 Confidence intervals and §3.6 Comparing mean values) |
|  | Lakens, D. (2013). [Calculating and reporting effect sizes to facilitate cumulative science: a practical primer for t-tests and ANOVAs](https://www.frontiersin.org/journals/psychology/articles/10.3389/fpsyg.2013.00863/full). *Frontiers in Psychology*, 4, 863. |
| **Practical** | *Statistical Thinking for the 21st Century* (Poldrack) — [Ch. 3, Summarizing data](https://statsthinking21.github.io/statsthinking21-core-site/summarizing-data.html) |

## Today's session

**Theory:** what a hypothesis test returns and what it does not; the sampling distribution and the standard error; the normal and t-distributions and degrees of freedom; confidence intervals and what the 95% guarantees; unstandardised and standardised effect sizes (and how *t* differs from Cohen's *d*); meta-analysis and selection for significance.

**Practical:** by the end you should be able to point R at a file with a relative path built by `here()`, manage packages with renv and load them, work with the t-distribution in R (density, critical values, probabilities), import a `.csv` file, wrangle it with `dplyr`, and compute confidence intervals and effect sizes, choosing the effect size that fits the question.

The session runs in three parts: a **joint walkthrough** (Part 1), **independent DataCamp work** (Part 2), and **in-notebook exercises** (Part 3). The **challenges** at the end are optional.

## Assessment

**Deadline for completion:** Sunday 11 October, 23:59.

1.  Complete the two assigned DataCamp courses ("Introduction to Importing Data in R" and "Data Manipulation with dplyr").
2.  Submit Worksheet 2 as a fully runnable `.qmd` file with the Part 3 exercises completed.
3.  Optional: the challenge exercises, and pushing your code to a GitHub repository (please still submit the `.qmd` file on Brightspace).

Name the file using your Student Number and surname as prefix, e.g. `2332334_Smith_worksheet_02.qmd`

Assignments 1 and 2 are mandatory and constitute 3% of your grade (pass/fail). Assignment 3 is optional and voluntary.

------------------------------------------------------------------------

## Part 1: Joint walkthrough

We will work through this section together. Run each chunk as we go and make sure you understand the output before moving on.

### Section A: Files, packages and paths

#### 1.1 Where is R looking? Working directories and paths

Every file on a computer has a **path**: the list of folders you pass through to reach it. When you ask R to read a file, R has to find it, and it starts looking in the **working directory**.

```{r}
#| eval: false
getwd()          # the folder R is currently working in
list.files()     # the files and folders inside it
```

There are two ways to write a path:

-   an **absolute path** starts from the top of the computer, e.g. `"/Users/ondrej/Documents/psy40760/seminar-02-worksheet_files/weight_trial.csv"` (Mac) or `"C:/Users/ondrej/Documents/psy40760/..."` (Windows);
-   a **relative path** starts from the working directory, e.g. `"seminar-02-worksheet_files/weight_trial.csv"`.

An absolute path only works on the computer it was written on: your colleague does not have a folder called `/Users/ondrej`. Relative paths travel with the project, which is what we want.

```{r}
#| eval: false
file.exists("seminar-02-worksheet_files/weight_trial.csv")   # relative path: TRUE if R can find it
```

Two special symbols appear in relative paths: `.` means "this folder" and `..` means "one folder up". So `"../data/file.csv"` means "go up one folder, then into `data`".

**The catch:** a relative path is only correct if the working directory is what you think it is. When you type in the console, the working directory is whatever you pointed R at (here, the course folder); when you run a chunk of a `.qmd`, or render it, it is the folder the `.qmd` sits in. If these differ, the same line of code finds the file in one case and fails in the other. You may be tempted to fix this with `setwd("/Users/...")`, but that puts an absolute path back into your code. Section 1.3 shows a better solution, which needs a package, so packages come first.

**Try it:** run `list.files("seminar-02-worksheet_files")`. What files are in this week's data folder?

```{r}
# your code here
```

#### 1.2 Packages: installing, loading and managing

Base R can do a lot, but most work in R relies on **packages**: bundles of functions written by others. Most are distributed through CRAN, R's official repository.

In this module we manage packages with **renv**. renv gives each project its own package library and records the exact version of every package in a file called `renv.lock`, so the code runs the same on your laptop, on ours, and in a year's time. The only time you use `install.packages()` is to install renv itself (you did this before Seminar 1):

```{r}
#| eval: false
install.packages("renv")   # once per computer; never needed again for other packages
```

There are two separate steps, and it helps to keep them apart:

| Step | Function | How often |
|----|----|----|
| **Install** (into the project library) | `renv::restore()` for the packages in `renv.lock`; `renv::install("pkg")` for anything else, then `renv::snapshot()` | when `renv.lock` changes, or when you add a package |
| **Load** (make available in this session) | `library(here)` | every session, at the top of your script |

This week we need `here`, `dplyr` and `tidyr`. After `git pull` and `renv::load()` (see "Before you arrive"), run, **in the console**:

```{r}
#| eval: false
renv::restore()                             # installs exactly the versions recorded in renv.lock
renv::install(c("here", "dplyr", "tidyr"))  # adds this week's packages (does nothing if already there)
renv::snapshot()                            # records them in your renv.lock
renv::status()                              # checks that the project library matches renv.lock
```

Run these in the console, not in the worksheet: an install line inside a `.qmd` would run again every time the document is rendered.

Check that a package is installed, and which version you have:

```{r}
requireNamespace("dplyr", quietly = TRUE)   # TRUE if installed
packageVersion("dplyr")
```

Now load the packages we use today:

```{r}
library(here)
library(dplyr)
```

When `dplyr` loads it prints a message that some objects are **masked**: `dplyr` has its own `filter()` and `lag()`, which now hide the versions in the `stats` package. You can always say exactly which one you mean with `package::function()`:

```{r}
#| eval: false
dplyr::filter(...)   # dplyr's filter: keep rows
stats::filter(...)   # stats' filter: a time-series operation
```

The `::` form also works without `library()`, which is handy when you need one function from a package once.

A few management tasks:

-   **Your own projects:** `renv::init()` sets up renv in a new project, `renv::install("pkg")` adds a package, and `renv::snapshot()` records the versions you use in `renv.lock`, so that `renv::restore()` can rebuild the same library on another computer.
-   **In the course folder** it works the same way: `renv::install("pkg")`, then `renv::snapshot()`. If `git pull` later complains that `renv.lock` has local changes, see the environment set-up guide.
-   `renv::update()` updates the project's packages and `renv::remove("name")` removes one.
-   `sessionInfo()` lists R's version and every loaded package: paste it into a help request so others can reproduce your setup.

**Try it:** run `renv::status()` and check which version of `here` you have installed.

```{r}
# your code here
```

#### 1.3 Relative paths with `here`

`here` solves the working-directory problem. It finds the **project root** and builds paths from there, whatever the working directory happens to be. In this course we mark the root of the practicals explicitly, with an empty file called `.root-practicals` (a hidden file: the name starts with a dot, so tick *Show Hidden Files* in the RStudio Files pane to see it). `here::i_am()` tells `here` to use the nearest folder, at or above the current one, that contains this file:

```{r}
here::i_am(".root-practicals")
here(".root-practicals")    # the practicals root on this computer
```

`here()` joins folder and file names into a full path for you. List the folders in order, then the file name:

```{r}
here("seminar-02-confidence-intervals-and-effect-sizes", "seminar-02-worksheet_files", "weight_trial.csv")
file.exists(here("seminar-02-confidence-intervals-and-effect-sizes", "seminar-02-worksheet_files", "weight_trial.csv"))
```

The code contains no absolute path, yet it gives R a complete one on every computer. The same line works when you run the chunk and when you render the document, on your laptop and on ours.

If `i_am()` gives an error, R could not find `.root-practicals` in the current folder or any folder above it: check with `getwd()` that your working directory is inside the course folder, and that `.root-practicals` is at its top (see "Before you arrive").

With a path in hand, reading the file is one line. You met `read.csv()` last week:

```{r}
trial <- read.csv(here("seminar-02-confidence-intervals-and-effect-sizes", "seminar-02-worksheet_files", "weight_trial.csv"))
head(trial)
```

We will come back to this data set in Part 3.

### Section B: The t-distribution and probabilities

With files and packages in place, we now turn to the statistics of the lecture: the t-distribution, critical values, and the probability of an event.

#### 1.4 From the normal distribution to the t-distribution

In the lecture, the sampling distribution of an estimate was approximated by the normal distribution, or by the t-distribution when the spread has to be estimated from the data. R has functions for both. We will build a picture of them in small steps.

**Step 1.** Make a grid of values along the x-axis.

```{r}
x <- seq(-4, 4, length.out = 200)
head(x)
```

**Step 2.** `dnorm()` gives the height of a normal curve at each value. A normal distribution has two **parameters**, its mean and its standard deviation, and `dnorm()` takes them as inputs: `dnorm(x, mean = ..., sd = ...)`. With `mean = 0` and `sd = 1` we get the **standard normal** distribution. Plot height against x as a line.

```{r}
plot(x, dnorm(x, mean = 0, sd = 1), type = "l", lwd = 2,
     xlab = "value", ylab = "density", main = "The standard normal distribution")
```

`mean = 0` and `sd = 1` are the defaults: if you leave them out, R uses these values, which is why the code below writes just `dnorm(x)`. Try changing them, e.g. `dnorm(x, mean = 1, sd = 2)`: the mean moves the curve along the x-axis, and the SD makes it wider and flatter.

**Step 3.** `dt()` does the same for the t-distribution. It has one parameter, the **degrees of freedom**, given as `df`: `dt(x, df = ...)`. The degrees of freedom count how many independent pieces of information are left for estimating the spread. With one sample of *n* people, one piece is used up by estimating the mean first, so `df = n - 1`; with two independent groups, one is used up per group mean, so `df = n1 + n2 - 2`. Unlike `mean` and `sd` in `dnorm()`, `df` has no default: you always have to give it. (There is no mean or SD to set: the t-distribution is already on a standardised scale, centred on 0.) Add a t-distribution with 3 df, i.e. a sample of 4 people, to the plot with `lines()`.

```{r}
plot(x, dnorm(x), type = "l", lwd = 2,
     xlab = "value", ylab = "density", main = "Normal vs t with 3 df")
lines(x, dt(x, df = 3), lwd = 2, col = "orange")
```

The t-distribution is lower in the middle and has **heavier tails**: extreme values are more likely. This is the price of estimating the spread from a small sample.

**Step 4.** Several degrees of freedom on one plot, with a legend.

```{r}
dfs  <- c(1, 3, 10, 30)
cols <- c("red", "orange", "darkgreen", "blue")

plot(x, dnorm(x), type = "l", lwd = 3, col = "black",
     xlab = "value", ylab = "density", main = "t-distributions with different df")
for (i in 1:4) {
  lines(x, dt(x, df = dfs[i]), lwd = 2, col = cols[i])
}
legend("topright", legend = c("normal", paste("t, df =", dfs)),
       col = c("black", cols), lwd = 2, bty = "n")
```

As the degrees of freedom grow, the t-distribution becomes indistinguishable from the normal. At 30 df the two lines almost overlap.

**Try it:** plot the normal curve together with a t-distribution with 100 df. Can you see any difference?

```{r}
# your code here
```

#### 1.5 The critical t-value

A 95% confidence interval uses the value that cuts off the middle 95% of the distribution: 2.5% in each tail. Such a value is a **quantile**, and the `q` functions return quantiles: you give them a probability, they return the value.

```{r}
qnorm(0.975)          # normal: 97.5% of the distribution lies below 1.96
qt(0.975, df = 5)     # t with 5 df: the cut-off is further out
```

Why 0.975 and not 0.95? We want 2.5% in the upper tail, so 97.5% lies below the cut-off.

The critical value shrinks towards 1.96 as the degrees of freedom grow:

```{r}
dfs <- c(2, 5, 10, 24, 46, 100, 1000)
round(qt(0.975, df = dfs), 3)
```

Trial A in the lecture had 24 + 24 − 2 = 46 df, hence its critical value of 2.013.

Now mark the critical values on the plot. `abline(v = ...)` draws vertical lines, and `polygon()` shades the two tails.

```{r}
df   <- 5
crit <- qt(0.975, df = df)

plot(x, dt(x, df = df), type = "l", lwd = 2,
     xlab = "t", ylab = "density", main = "t with 5 df: the middle 95% and the two 2.5% tails")
abline(v = c(-crit, crit), lty = 2, col = "red")

upper <- x[x >= crit]
polygon(c(crit, upper, max(upper)), c(0, dt(upper, df), 0), col = "red", border = NA)
lower <- x[x <= -crit]
polygon(c(min(lower), lower, -crit), c(0, dt(lower, df), 0), col = "red", border = NA)

text(crit, 0.3, paste("t =", round(crit, 3)), pos = 4)
```

**Try it:** find the critical value for a **99%** interval with 20 df. Is it larger or smaller than the 95% one, and why?

```{r}
# your code here
```

#### 1.6 Probabilities: the density and the cumulative distribution

For a continuous variable, the height of the curve (the **density**) is not a probability. Probabilities are **areas** under the curve. Two functions answer two different questions:

| Function | Question it answers |
|----|----|
| `dt(q, df)` | How high is the curve at `q`? (a density, not a probability) |
| `pt(q, df)` | What is the probability of a value **at or below** `q`? (the area to the left) |

```{r}
dt(1.5, df = 10)   # height of the curve at 1.5
pt(1.5, df = 10)   # P(T <= 1.5)
```

The probability of any **single exact value**, such as exactly 1.5, is zero: a line has no area. We can only ask about ranges of values.

`pt()` is the **cumulative distribution function**: it adds up the area from the far left up to `q`. Plotting it against `q` gives an S-shaped curve that rises from 0 to 1.

```{r}
df <- 10
par(mfrow = c(1, 2))   # two plots side by side

plot(x, dt(x, df), type = "l", lwd = 2, xlab = "t", ylab = "density", main = "Density: dt()")
left <- x[x <= 1.5]
polygon(c(min(left), left, 1.5), c(0, dt(left, df), 0), col = "lightblue", border = NA)
lines(x, dt(x, df), lwd = 2)

plot(x, pt(x, df), type = "l", lwd = 2, xlab = "t", ylab = "P(T <= t)", main = "Cumulative: pt()")
points(1.5, pt(1.5, df), pch = 19, col = "blue", cex = 1.5)
abline(h = pt(1.5, df), lty = 2, col = "blue")

par(mfrow = c(1, 1))   # back to one plot at a time
```

The shaded **area** on the left and the **height** of the point on the right are the same number, 0.918.

The four functions for each distribution follow one naming pattern. For the t-distribution:

| Prefix | Function | Gives |
|----|----|----|
| `d` | `dt()` | density (height of the curve) |
| `p` | `pt()` | probability of a value at or below `q` |
| `q` | `qt()` | the value below which a given probability lies (the inverse of `pt()`) |
| `r` | `rt()` | random draws from the distribution |

The same pattern holds for the normal (`dnorm`, `pnorm`, `qnorm`, `rnorm`) and for other distributions.

**Try it:** check that `qt()` undoes `pt()`: compute `pt(1.5, df = 10)`, then feed the result into `qt()` with the same df.

```{r}
# your code here
```

#### 1.7 Bigger than, smaller than, and between

`pt()` always gives the area to the **left**. Every other question can be built from it.

**Smaller than** a value:

```{r}
pt(2, df = 10)                       # P(T <= 2)
```

**Bigger than** a value: the total area is 1, so subtract.

```{r}
1 - pt(2, df = 10)                   # P(T > 2)
pt(2, df = 10, lower.tail = FALSE)   # the same, asked directly
```

**Between** two values: the area to the left of the upper value, minus the area to the left of the lower value.

```{r}
pt(2, df = 10) - pt(-1, df = 10)     # P(-1 < T < 2)
```

**More extreme than** a value in either direction. This is the **two-sided p-value**: the probability of a t at least this far from 0, if the null hypothesis were true. Because the distribution is symmetric, double one tail.

```{r}
2 * pt(-abs(2), df = 10)             # P(|T| >= 2)
```

Check this against Trial A from the lecture: t(46) = 3.03, p = .004.

```{r}
2 * pt(-3.03, df = 46)
```

The same reasoning works for any normal distribution. `pnorm()` takes the mean and SD directly. Suppose weight change over 12 weeks in a control group is normally distributed with mean 1 kg and SD 5 kg. What proportion of people lose more than 5 kg?

```{r}
1 - pnorm(5, mean = 1, sd = 5)               # P(loss > 5)
pnorm(2, mean = 1, sd = 5) - pnorm(0, mean = 1, sd = 5)   # P(0 < loss < 2)
```

**Try it:** with the same distribution (mean 1 kg, SD 5 kg), what proportion of people **gain** weight, i.e. have a weight loss below 0?

```{r}
# your code here
```

------------------------------------------------------------------------

## Part 2: DataCamp

Complete the following two assigned courses on DataCamp. They count towards your grade (see Assessment above).

-   [**Introduction to Importing Data in R**](https://app.datacamp.com/learn/courses/introduction-to-importing-data-in-r): reading flat files with `read.csv()`, `read.delim()` and `read.table()`, file paths, column names and column classes.
-   [**Data Manipulation with dplyr**](https://app.datacamp.com/learn/courses/data-manipulation-with-dplyr): `select()`, `filter()`, `arrange()`, `mutate()`, `count()`, `group_by()` + `summarize()`, and chaining them with the pipe `%>%`.

Part 3 uses both, so it is worth doing the DataCamp courses first.

------------------------------------------------------------------------

## Part 3: Exercises in RStudio

These exercises are completed independently, in this notebook. All 30 are mandatory. Write your code in the chunks provided and check each result before moving on. Where a question asks you to explain or interpret, write your answer as a comment (`#`) in the chunk, or as text below it.

**The data.** `weight_trial.csv` comes from a fictional (simulated) 12-week trial of a weight-loss programme. Participants were randomised to the **programme** or to a **control** group at one of three sites. The columns are:

| Column | Meaning |
|----|----|
| `id` | participant ID |
| `site` | recruitment site: Dublin, Cork or Galway |
| `group` | `control` or `programme` |
| `sex`, `age` | participant's sex and age (years) |
| `weight_start_kg` | weight at the start (kg) |
| `weight_end_kg` | weight after 12 weeks (kg) |
| `wellbeing` | score on a wellbeing questionnaire at 12 weeks (0–40; higher = better) |

As in the lecture, assume that a weight loss is only worth the cost of the programme if it is at least **2 kg** (the smallest effect of interest).

We use the pipe `%>%` from `dplyr`, as in DataCamp. (R also has a built-in pipe, `|>`, which works the same way for everything we do here.)

### Importing the data (1–5)

### Exercise 1 — Packages

Load the `here` and `dplyr` packages.

```{r}
# your code here
```

### Exercise 2 — Building the path

Use `here()` to build the path to `weight_trial.csv` and store it in a variable called `data_path`. Use `file.exists()` to check that R can find the file.

```{r}
# your code here
```

### Exercise 3 — Importing

Import the file with `read.csv()` and store it as `trial`. How many rows and columns does it have?

```{r}
# your code here
```

### Exercise 4 — Checking what was imported

Use `str()` (or `glimpse()` from `dplyr`) to look at the column types. Which columns are character, which are numeric? Then use `count()` to find how many participants are in each group.

```{r}
# your code here
```

### Exercise 5 — Missing values

How many missing values (`NA`) does each column contain? *Hint:* `is.na()` returns `TRUE`/`FALSE` for every value, and `colSums()` adds these up column by column. Which participants have a missing `wellbeing` score?

```{r}
# your code here
```

### Wrangling (6–12)

### Exercise 6 — `mutate()`: the outcome

Add a column `weight_loss_kg` equal to the start weight minus the end weight (so that a positive number means weight was lost). Save the result back into `trial`.

```{r}
# your code here
```

### Exercise 7 — `select()` and `relocate()`

Create a smaller data frame `trial_small` with only `id`, `site`, `group` and `weight_loss_kg`, in that order. Then, in the full `trial` data, move `weight_loss_kg` so that it is the column directly after `group` (*hint:* `relocate()` with `.after =`).

```{r}
# your code here
```

### Exercise 8 — `filter()`

How many programme participants lost at least 2 kg? How many control participants did?

```{r}
# your code here
```

### Exercise 9 — `arrange()` and `slice_max()`

Show the five participants who lost the most weight, with the largest loss first. Which groups are they in?

```{r}
# your code here
```

### Exercise 10 — `group_by()` and `summarise()`

For each group, compute the number of participants, the mean and the SD of `weight_loss_kg`. Store the result as `loss_summary`; you will need these numbers later.

```{r}
# your code here
```

### Exercise 11 — Counting combinations

Use `count()` to show how many participants there are in each combination of `site` and `group`. Are the sites the same size?

```{r}
# your code here
```

### Exercise 12 — Recoding with `ifelse()`

Add a column `worthwhile` that is `"yes"` if a participant lost at least 2 kg and `"no"` otherwise. Then compute the **proportion** of `"yes"` in each group. *Hint:* `mean(worthwhile == "yes")` inside `summarise()`.

```{r}
# your code here
```

### Probabilities (13–18)

### Exercise 13 — A normal model for the control group

Suppose weight loss in the control **population** is normally distributed, with the mean and SD you found for the control group in Exercise 10. Under this model, what is the probability that a control participant loses more than 5 kg? Compare this with the proportion of control participants in the data who actually lost more than 5 kg.

```{r}
# your code here
```

### Exercise 14 — Between two values

Under the same model, what is the probability that a control participant's weight change lies between −2 kg and +2 kg?

```{r}
# your code here
```

### Exercise 15 — Smaller than, for the programme group

Now build the same kind of normal model for the **programme** group. What proportion of programme participants would be expected to lose **less** than the 2 kg threshold?

```{r}
# your code here
```

### Exercise 16 — Critical values

For a t-distribution with the degrees of freedom of the programme group alone (n − 1), find the critical values for a 90%, a 95% and a 99% interval. Put them in one vector.

```{r}
# your code here
```

### Exercise 17 — A t statistic by hand

For the programme group only, test whether the mean weight loss differs from 0. Compute by hand: the mean, the SD, n, the standard error (SE = SD / √n), the t statistic (t = mean / SE) and the two-sided p-value with `pt()`. Then check your t and p against `t.test(x)`.

```{r}
# your code here
```

### Exercise 18 — Where does a value lie?

Using your normal model for the programme group (Exercise 15), which weight loss is exceeded by only 10% of programme participants? *Hint:* this is a quantile, so use `qnorm()`. Compare it with the empirical value from `quantile(x, 0.9)`.

```{r}
# your code here
```

### Confidence intervals (19–24)

### Exercise 19 — A CI for one mean, by hand

Compute the 95% confidence interval for the mean weight loss in the **programme** group by hand: mean ± t~crit~ × SE, with t~crit~ from `qt()` and n − 1 degrees of freedom.

```{r}
# your code here
```

### Exercise 20 — The same CI from `t.test()`

Run `t.test()` on the programme group's weight loss and extract the confidence interval with `$conf.int`. Does it match Exercise 19?

```{r}
# your code here
```

### Exercise 21 — Changing the confidence level

Compute the 90% and the 99% intervals for the same mean (the `conf.level` argument of `t.test()`). Which interval is widest? Explain in one sentence why.

```{r}
# your code here
```

### Exercise 22 — A CI for the difference between groups, by hand

Compute the 95% CI for the **difference** in mean weight loss (programme minus control) by hand, using

$$SE_{diff} = \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}$$

and a critical value with n₁ + n₂ − 2 degrees of freedom.

```{r}
# your code here
```

### Exercise 23 — The same CI from `t.test()`

Run `t.test(weight_loss_kg ~ group, data = trial)`. Extract the CI, the t statistic and the p-value. Two things to notice: (a) the sign — R subtracts the groups in alphabetical order, so this is control **minus** programme; (b) the interval differs slightly from yours, because `t.test()` uses the Welch correction for unequal variances, which changes the degrees of freedom. What are its degrees of freedom?

```{r}
# your code here
```

### Exercise 24 — Reading the interval

Compare your interval for the difference with 0 and with the 2 kg threshold. Is the difference statistically significant at the 5% level? Can we conclude that the programme produces a worthwhile loss? Which of the lecture's two problems, **precision** or **relevance**, does this trial have? Answer in two or three sentences.

```{r}
# your answer here (as comments)
```

### Effect sizes (25–30)

### Exercise 25 — An unstandardised effect size

Report the effect of the programme on weight as an unstandardised effect size: the mean difference in kg with its 95% CI. Write the one-sentence result as you would in a paper.

```{r}
# your code here
```

### Exercise 26 — Cohen's *d* by hand

Compute Cohen's *d* for weight loss: the mean difference divided by the pooled SD,

$$d = \frac{M_1 - M_2}{SD_{pooled}}, \qquad SD_{pooled} = \sqrt{\frac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}{n_1 + n_2 - 2}}$$

```{r}
# your code here
```

### Exercise 27 — Wellbeing, with missing values

Compute the mean difference and Cohen's *d* for `wellbeing` (programme minus control). Remember that four scores are missing: remove them first (e.g. `filter(!is.na(wellbeing))`), and make sure the *n* in each group is the number of non-missing scores.

```{r}
# your code here
```

### Exercise 28 — Standardised or unstandardised?

For each case below, decide whether you would report an **unstandardised** or a **standardised** effect size, and justify your choice in one sentence (see the lecture and Lakens, 2013):

(a) the effect of the programme on weight loss, reported to a GP deciding whether to recommend it;
(b) the effect of the programme on wellbeing (a 0–40 questionnaire);
(c) comparing this trial's wellbeing effect with another trial that measured wellbeing on a different, 1–7 scale;
(d) a reaction-time study reporting how much faster caffeine drinkers are.

Then report the effect for (a) and (b) accordingly, using your results from Exercises 25–27.

```{r}
# your answers here (as comments), plus any code
```

### Exercise 29 — *t* and *d* are not the same number

For two groups of equal size *n*, t ≈ d × √(n/2). Check this with your *d* from Exercise 26 and the group size. Then answer: if the trial had recruited four times as many participants and found the same means and SDs, what would happen to *d*, to *t* and to the width of the CI?

```{r}
# your code here
```

### Exercise 30 — Site by site

Compute the difference in mean weight loss (programme minus control) and its 95% CI **separately for each site**. *Hint:* `filter()` one site at a time and run `t.test()` on it. Which site gives the widest interval, and why? Does any single site on its own show a significant effect? What does this suggest about analysing small subgroups?

```{r}
# your code here
```

------------------------------------------------------------------------

## Challenge exercises (optional)

**These challenges are optional and do not count towards the pass mark.** They are harder and go beyond what we covered together. Try them if you finish early, or come back to them later in the term.

They use a second data set, `weight_followup_wide.csv`: a smaller follow-up study (40 people, programme or control) in which each person was weighed four times, at weeks 0, 4, 8 and 12. The data are in **wide format**: one row per person, one column per measurement (`weight_wk00` … `weight_wk12`). Most analyses need **long format**: one row per measurement, with a column saying which week it is. Reshaping uses the `tidyr` package.

```{r}
#| eval: false
library(tidyr)
```

### Challenge 1 — From wide to long

Import `weight_followup_wide.csv` with `here()` and `read.csv()`. Use `pivot_longer()` to turn it into a long data frame `followup_long` with columns `id`, `group`, `week` and `weight_kg`. Make `week` a number (0, 4, 8, 12), not text. *Hint:* look at the `names_to`, `names_prefix` and `values_to` arguments, and `as.numeric()`. How many rows does the long data frame have, and why?

```{r}
# your code here
```

### Challenge 2 — Summaries and a plot

From `followup_long`, compute the mean weight for each group at each week. Plot the two groups' means against week in one base-R plot (`plot()` for the first group, `lines()` for the second, and a legend).

```{r}
# your code here
```

### Challenge 3 — Change from baseline

Compute each person's change from week 0 to week 12. (You can do this in the wide data directly, or go back from long to wide with `pivot_wider()`.) Then give the difference between groups in mean change, its 95% CI, and Cohen's *d*. Compare with the main trial: which study estimates the effect more precisely, and why?

```{r}
# your code here
```

### Challenge 4 — Within-subject effect sizes

For the **programme** group only, compare week 0 with week 12 within the same people. Run a paired t-test (`t.test(x, y, paired = TRUE)`). Then compute two versions of Cohen's *d* (Lakens, 2013):

-   *d*~z~ = mean of the differences / SD of the differences;
-   *d*~av~ = mean of the differences / average of the SDs at week 0 and week 12.

Why are they so different? Which would you report, and when?

```{r}
# your code here
```

### Challenge 5 — Writing functions and testing the 95%

(a) Write a function `ci_mean(x, level = 0.95)` that returns the confidence interval for the mean of `x`, and a function `cohens_d(x, y)` that returns Cohen's *d* with the pooled SD. Check both against your earlier answers.

(b) Use `ci_mean()` to test what "95%" means. Draw 1,000 samples of size 10 from a normal distribution with mean 2 and SD 5 (`rnorm()`), compute the interval for each, and count how often it contains the true mean of 2. Repeat with samples of size 100. What changes, and what stays the same?

```{r}
# your code here
```
